---
title: "Sleep duration, mortality and the influence of age"
date: 30/08/2017
author:
  - name: Rino Bellocco
    orcid: 0000-0002-2538-0287
    affiliation: Department of Statistics and Quantitative Methods, University of Milano-Bicocca, Milan, Italy
pub-info:
  reference: >-
    Rino Bellocco, “Sleep duration, mortality and the influence of age” <em>EJE</em> 32, (August 2017): 881–891, doi: <a href="https://doi.org/10.1007/s10654-017-0297-0"><code>10.1007/s10654-017-0297-0</code></a>
  links:
  - name: Article
    url: https://doi.org/10.1007/s10654-017-0297-0
    icon: fa-solid fa-scroll
  - name: Add to Zotero
    url: https://www.zotero.org/save?type=doi&q=10.1007/s10654-017-0297-0
    icon: ai ai-zotero
doi: 10.1007/s10654-017-0297-0
---

## Important links

- [Article](https://doi.org/10.1007/s10654-017-0297-0)

## Abstract

Prior work has shown that both short and long sleep predict mortality. However, sleep duration decreases with age and this may affect the relationship of sleep duration with mortality. The purpose of the present study was to assess whether the association between sleep duration and mortality varies with age. Prospective cohort study. 43,863 individuals (64% women), recruited in September 1997 during the Swedish National March and followed through record-linkages for 13 years. Sleep duration was self-reported and measured using the Karolinska Sleep Questionnaire, and grouped into 4 categories: ≤5, 6, 7 (reference) and ≥8 h. Up to 2010 3548 deaths occurred. Multivariable Cox proportional hazards regression models with attained age as time scale were fitted to estimate mortality rate ratios. Among individuals <65 years, short (≤5 h) and long (≥8 h) sleep duration showed a significant relationship with mortality (HR 1.37, 95% CI 1.09–1.71, and HR 1.27, 95% CI 1.08–1.48). Among individuals 65 years or older, no relationships between sleep duration and mortality were observed. The effect of short and long sleep duration on mortality was highest among young individuals and decreased with increasing age. The results suggest that age plays an important role in the relationship between sleep duration and mortality.

## DAG

```{r}
dag <- dagitty::dagitty("dag {
  y <- x <- z1 <- v -> z2 -> y
  z1 <- w1 <-> w2 -> z2
  x <- w1 -> y
  x <- w2 -> y
  x [exposure]
  y [outcome]
}")
tidy_dag <- ggdag::tidy_dagitty(dag)

ggdag::ggdag_status(tidy_dag) +
  ggdag::theme_dag()
```

The **adjustment sets** are:

```{r}
dagitty::adjustmentSets(dag)
```

## Citation

<p class="buttons" style="text-align:center;">
<a class="btn btn-danger" target="_blank" href="https://www.zotero.org/save?type=doi&q=10.1007/s10654-017-0297-0">{{< ai zotero >}}&emsp;Add to Zotero {{< fa square-arrow-up-right >}}</a>
</p>

```bibtex
@article{aakerstedt2017sleep,
  title={Sleep duration, mortality and the influence of age},
  author={{\AA}kerstedt, Torbj{\"o}rn and Ghilotti, Francesca and Grotta, Alessandra and Bellavia, Andrea and Lagerros, Ylva Trolle and Bellocco, Rino},
  journal={European journal of epidemiology},
  volume={32},
  pages={881--891},
  year={2017},
  publisher={Springer}
}
```
