---
title: "The SSQ model of schizophrenic prodromal unfolding revised: An analysis of its causal chains based on the language of directed graphs"
date: 09/01/2014
author:
  - name: D van Kampen
    affiliation: Department of Clinical Psychology, Vrije Universiteit, Amsterdam, The Netherlands
pub-info:
  reference: >-
    D van Kampen, “The SSQ model of schizophrenic prodromal unfolding revised: An analysis of its causal chains based on the language of directed graphs” <em>Eur Psychiatry</em> 29, 7 (September 2014): 437-448, doi: <a href="https://doi.org/10.1016/j.eurpsy.2013.11.001"><code>10.1016/j.eurpsy.2013.11.001</code></a>
  links:
  - name: Article
    url: https://doi.org/10.1016/j.eurpsy.2013.11.001
    icon: fa-solid fa-scroll
  - name: Add to Zotero
    url: https://www.zotero.org/save?type=doi&q=10.1016/j.eurpsy.2013.11.001
    icon: ai ai-zotero
doi: 10.1016/j.eurpsy.2013.11.001
---

## Important links

- [Article](https://doi.org/10.1016/j.eurpsy.2013.11.001)

## Abstract

### Objective
Except in broad outline, little is known about the most likely symptomatic trajectories in prodromal schizophrenia. The aim of this study is to delineate these pathways.

### Methods
Taking into account existing clinical knowledge, the causal relationships between the 12 prodrome scales of the Schizotypic Syndrome Questionnaire (SSQ) were examined in a general-population sample by applying the mathematical theory of directed graphs. Use was made of two discovery algorithms implemented in the Tetrad-4 program, as well as of the graphical DAGitty program to test whether a particular model holds.

### Results
A promising model was selected that may describe the causal pathways in schizophrenic prodromal unfolding. Testing this model by means of DAGitty, it was shown that the minimal testable implications, listed as conditional independences, and the direct and total effects in the model, identified after correction for bias, were as hypothesized. For practical reasons, a simpler version of the resulting SSQ model, containing only its principal pathways, was provided.

### Conclusions
Although resembling an earlier model that was based on a series of LISREL analyses, the present model was believed to provide a more dependable description of schizophrenic prodromal unfolding, as it relies on methods that are less subject to the limitations involved in SEM.

## DAG

```{r}
dag <- dagitty::dagitty('dag {
AFF [pos="0.262,0.477"]
AIS [pos="0.123,0.736"]
ALN [pos="0.438,0.506"]
APA [pos="0.376,0.147"]
CDR [pos="0.628,0.332"]
DET [pos="0.920,0.561"]
EGC [outcome,pos="0.916,1.016"]
FTW [pos="0.667,0.639"]
HOS [pos="0.886,0.727"]
PER [pos="0.920,0.382"]
SAN [pos="0.031,0.371"]
SUS [exposure,pos="0.295,1.003"]
AFF -> ALN
AFF -> APA
AFF -> CDR
AIS -> AFF
AIS -> EGC
AIS -> SUS
ALN -> APA
ALN -> DET
ALN -> FTW
ALN -> PER
ALN -> SUS
CDR -> DET
EGC -> HOS
FTW -> DET
FTW -> EGC
PER -> DET
SAN -> AFF
SAN -> AIS
SAN -> ALN
SAN -> APA
SAN -> CDR
SUS -> EGC
SUS -> FTW
SUS -> HOS
}')
tidy_dag <- ggdag::tidy_dagitty(dag)

ggdag::ggdag_status(tidy_dag) +
  ggdag::theme_dag()
```

The **adjustment sets** are:

```{r}
dagitty::adjustmentSets(dag)
```

## Citation

<p class="buttons" style="text-align:center;">
<a class="btn btn-danger" target="_blank" href="https://www.zotero.org/save?type=doi&q=10.1016/j.eurpsy.2013.11.001">{{< ai zotero >}}&emsp;Add to Zotero {{< fa square-arrow-up-right >}}</a>
</p>

```bibtex
@article{van2014ssq,
  title={The SSQ model of schizophrenic prodromal unfolding revised: An analysis of its causal chains based on the language of directed graphs},
  author={van Kampen, Dirk},
  journal={European Psychiatry},
  volume={29},
  number={7},
  pages={437--448},
  year={2014},
  publisher={Cambridge University Press}
}
```
